Estimating Animal Abundance with N-Mixture Models Using the \pkg{R-INLA} Package for R

### 2017-05-09 meehan_et_al_inla_nmix_jss/article.tex
more tweaks
Minor edit of the R code in the Appendix

%% set up %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[article]{jss}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{natbib}

@@

The key observation is that both the Poisson and the Binomial distribution can be evaluated recursively in $n$,
\begin{displaymath}
    \text{Poisson}(n; \lambda) = \text{Poisson}(n-1; \lambda) \frac{\lambda}{n}
\end{displaymath}
and
\begin{displaymath}
    \text{Binomial}(y; n, p) = \text{Binomial}(y; n-1, p) \frac{n}{n-y}(1-p),
\end{displaymath}
and then also for the Poisson-Binomial product
\begin{displaymath}
    \text{Poisson}(n ; \lambda) \; \text{Binomial}(y;  n, p)
    =
    \text{Poisson}(n-1; \lambda) \; \text{Binomial}(y; n-1, p)
    \frac{\lambda}{n-y}(1-p).
\end{displaymath}
If we define $f_i = \lambda(1-p)/i$ for $i=1, 2, \ldots$, we can make use of this recursive form to express the likelihood with a finite upper limit as
\begin{eqnarray}
    \text{Probability}(y) &=& \sum_{n = y}^{n_{\text{max}}}
                       \text{Poisson}(n ; \lambda)\;
                      \text{Binomial}(y;  n, p) \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{Binomial}(y; y, p)
                       \Big\{ 1 + f_1 + f_1f_2 +
                    \ldots
                      +f_1\cdots f_{n_\text{max}}
                       \Big\} \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{Binomial}(y; y, p)
                       \Big\{ 1 + f_1(1+f_2(1+f_3(1+ \dots)))\Big\}\nonumber
\end{eqnarray}
The log-likelihood can then be evaluated using the following simple \proglang{R}-code.
\begin{CodeInput}
R> fac <- 1; ff <- lambda * (1-p)
R> for(i in (n.max - y):1){
+    fac <- 1 + fac * ff / i
+  } 
R> log.L <- dpois(y, lambda, log = TRUE) + dbinom(y, y, p, log = TRUE) + 
+           log(fac)
\end{CodeInput}
Since this evaluation is recursive in decreasing $n$, we have to choose the upper limit $n_\text{max}$ up-front, for example as an integer larger than $y$ so that $\frac{\lambda (1-p)}{n_\text{max}-y}$ is small. Note that we are computing \code{fac} starting with the smallest contributions, which are more numerically stable.

>>>>

%% set up %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[article]{jss}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{hyperref}

@@

The key observation is that both the Poisson and the Binomial distribution can be evaluated recursively in $n$,
\begin{displaymath}
    \text{Poisson}(n; \lambda) = \text{Poisson}(n-1; \lambda) \frac{\lambda}{n}
\end{displaymath}
and
\begin{displaymath}
    \text{Binomial}(y; n, p) = \text{Binomial}(y; n-1, p) \frac{n}{n-y}(1-p),
\end{displaymath}
and then also for the Poisson-Binomial product
\begin{displaymath}
    \text{Poisson}(n ; \lambda) \; \text{Binomial}(y;  n, p)
    =
    \text{Poisson}(n-1; \lambda) \; \text{Binomial}(y; n-1, p)
    \frac{\lambda}{n-y}(1-p).
\end{displaymath}
If we define $f_i = \lambda(1-p)/i$ for $i=1, 2, \ldots$, we can make use of this recursive form to express the likelihood with a finite upper limit as
\begin{eqnarray}
    \text{Probability}(y) &=& \sum_{n = y}^{n_{\text{max}}}
                       \text{Poisson}(n ; \lambda)\;
                      \text{Binomial}(y;  n, p) \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{Binomial}(y; y, p)
                       \Big\{ 1 + f_1 + f_1f_2 +
                    \ldots
                      +f_1\cdots f_{n_\text{max}}
                       \Big\} \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{Binomial}(y; y, p)
                       \Big\{ 1 + f_1(1+f_2(1+f_3(1+ \dots)))\Big\}\nonumber
\end{eqnarray}
The log-likelihood can then be evaluated using the following simple \proglang{R}-code.
\begin{CodeInput}
R> fac <- 1; ff <- lambda * (1-p)
R> for(i in (n.max - y):1) fac <- 1 + fac * ff / i
R> log.L <- dpois(y, lambda, log = TRUE) +
+           dbinom(y, y, p, log = TRUE) + log(fac)
\end{CodeInput}
Since this evaluation is recursive in decreasing $n$, we have to choose the upper limit $n_\text{max}$ up-front, for example as an integer larger than $y$ so that $\frac{\lambda (1-p)}{n_\text{max}-y}$ is small. Note that we are computing \code{fac} starting with the smallest contributions, which are more numerically stable.
